### 2020-01-28 paper/sex_selection_spacing-ver3.tex
Minor edits and additions throughout paper

\date{December 2018}

@@

What the prior research on India has not appreciated is that since each abortion 
increases the interval between births by six to twelve months, the growing use of 
sex-selective abortions has the potential to increase birth spacing substantially. %
\footnote{
The increase consists of three parts.
First, starting from the abortion, the uterus needs at 
least two menstrual cycles to recover;  otherwise, the likelihood 
of spontaneous abortion increases substantially \citep{zhou00b}.
The second part is the waiting time to conception, which is between 
one and six months \citep{Wang2003}.
Finally, sex determination tests are reliable only from three months 
of gestation onwards.
}
Add to this increase that the combination of increasing female education, 
lower fertility, higher household income, and the low and declining female 
labor force participation are also likely to impact birth spacing.
Hence, the changes in birth spacing may be much more substantial in India 
than what we have observed in other countries as they developed.

@@

First, tempo effects, arising from changes in birth spacing, bias the total 
fertility rate (TFR), in the sense that the TFR becomes a more biased measure 
of cohort fertility \citep{Bongaarts1999,Ni-Bhrolchain2011}.
Hence, fertility may be, in fact, substantially higher than what is generally 
accepted based on the TFR.

@@

Finally, birth spacing has long served as a measure of son preference, with strong
son preference associated with shorter birth intervals, the fewer sons a family has 
\citep{ben-porath76b,Leung1988}.
Now, longer birth intervals in the absence of sons, rather than an indicator of 
lower son preference, may instead arise because couples with a strong preference 
for sons use sex selection, possibly to the point where we observe a reversal of 
the traditional spacing pattern.
Families with son preference who---for one reason or another---do not use prenatal 
sex selection likely continue to have shorter birth spacing.

Here, I examine how birth spacing in India has changed over time and across 
groups with the introduction of sex selection.
I apply a competing risk hazard model that directly incorporates the effects of 
sex-selective abortions on birth intervals
\emph{and} 
the likelihood of a son, 
but which also works in cases without prenatal sex selection.
I apply the method to birth histories of Hindu women, using data from the four
India's National Family and Health Surveys (NFHS), covering the period 
1972 to 2016.

% [Hypotheses / questions]
I address four main questions.
[TK combine questions and highlight of answers here?]

@@

The standard approach in the birth spacing literature is to use proportional hazard
models with a single exit, the birth of a child.
There are two problems with the standard approach in this setting.
First, and most importantly, the introduction of sex selection means that the sex of the 
next child is no longer necessarily random and that parents' choices will impact the 
spacing to the birth of a girl or a boy differently.
I, therefore, use a competing risk set-up, which can capture both the non-randomness of
the birth outcome and the differential spacing.
To my knowledge, this is the first application of competing risk models to birth
spacing.%
\footnote{
\cite{Merli2000} used a discrete hazard model to examine whether 
there were under-reporting of births in rural China, although they 
estimated separate waiting time regressions for boys and girls.
}

@@

The TFR measure uses births up to parity four to make the measure comparable to what I 
can calculate for the hazard model, and I calculate this four-parity TFR by urban/rural 
and the education groups used above.
Hence, the TFRs are not directly comparable to those in the NFHS reports, although the
differences should be smaller the higher the education group and the most recent the 
survey provided that cohort fertility has fallen over time.
Otherwise, I follow the procedure used in the NFHS reports and calculate age specific 
fertility rates for five-year age groups using the number of births in months 1 to 36 
before the survey month (births in the survey month are not included) and the number
of women in each age group.%
\footnote{
See \citet{Croft2018} for more detail.
Because NFHS-1 and NFHS-2 did not interview all women about fertility, the number of women 
is based on the household rosters, assuming that never-married women have had no children.
Because of the low number of births to women aged 45 to 49 this age group was not
used for the hazard model estimations and, therefore, is also excluded for the fertility
rate.
}

@@

The central question addressed here is the extent to which spacing patterns significantly 
changed as prenatal sex determination became available.
The underlying idea is that in the absence of sex selection son preference leads to shorter 
spacing after the birth of a girl than after the birth of a boy, whereas when prenatal sex 
selection is available son preference increases the spacing after the birth of a girl 
relative to after the birth of a boy.
I introduce an empirical method that simultaneously accounts for spacing between births 
and the potential use of sex selection, and apply the method to four decades of data 
from India's NFHS.

>>>>

\date{January 2019}

@@

What the prior research on India has not appreciated is that since each abortion 
increases the interval between births by six to twelve months, the growing use of 
sex-selective abortions has the potential to substantially increase birth spacing.%
\footnote{
The increase consists of three parts.
First, starting from the abortion, the uterus needs at 
least two menstrual cycles to recover;  otherwise, the likelihood 
of spontaneous abortion increases substantially \citep{zhou00b}.
The second part is the waiting time to conception, which is between 
one and six months \citep{Wang2003}.
Finally, sex determination tests are reliable only from three months 
of gestation onwards.
}
Add to this increase that the combination of increasing female education, 
lower fertility, higher household income, and the low and declining female 
labor force participation are also likely to impact birth spacing.
Hence, the changes in birth spacing may be much more substantial in India 
than what we have observed in other countries as they developed.

@@

First, tempo effects, arising from changes in birth spacing, bias the total 
fertility rate (TFR), in the sense that the TFR becomes a more biased measure 
of cohort fertility \citep{Hotz1997,Bongaarts1999,Ni-Bhrolchain2011}.
If birth spacing increased substantially over time, India's fertility may, in 
fact, be substantially higher than what is generally accepted based on the TFR.

@@

Finally, birth spacing has long served as a measure of son preference, with strong
son preference associated with shorter birth intervals, the fewer sons a family has 
\citep{ben-porath76b,Leung1988}.
Now, longer birth intervals in the absence of sons, rather than an indicator of 
lower son preference, may instead arise because couples with a strong preference 
for sons use sex selection, possibly to the point where we observe a reversal of 
the traditional spacing pattern.
Despite a large literature on son preference and birth spacing, there is, however,
no prior research has examined how sex selection affects spacing.

Here, I examine how birth spacing in India has changed over time and across 
groups with the introduction of sex selection.
Using data from the four National Family and Health Surveys (NFHS), I apply
a competing risk hazard model to the birth histories of Hindu women, covering
the period 1972 to 2016.
The competing risk hazard model directly incorporates the effects of 
sex-selective abortions on birth intervals
\emph{and} 
the likelihood of a son, 
but also works in cases without prenatal sex selection.

% [Hypotheses / questions]
I address four main questions.

@@

The standard approach in the birth spacing literature is to use proportional hazard
models with a single exit, the birth of a child.%
\footnote{
See \citet{Sheps1970} and \citet{Newman1984} for early discussions of why 
hazard models are the preferred way to deal with the censoring of birth
intervals.
}
There are, however, two problems with the standard approach in this setting.
First, and most importantly, the introduction of sex selection means that the sex of the 
next child is no longer necessarily random and that parents' choices will impact the 
spacing to the birth of a girl or a boy differently.
I, therefore, use a competing risk set-up, which can capture both the non-randomness of
the birth outcome and the differential spacing.
To my knowledge, this is the first application of competing risk models to birth
spacing.%
\footnote{
\cite{Merli2000} used a discrete hazard model to examine whether 
there were under-reporting of births in rural China, although they 
estimated separate waiting time regressions for boys and girls.
}

@@

The TFR measure uses births up to parity four to make the measure comparable to what I 
can calculate for the hazard model, and I calculate this four-parity TFR by urban/rural 
and the education groups used above.
Hence, the TFRs are not directly comparable to those in the NFHS reports, although the
differences should be smaller the higher the education group and the most recent the 
survey provided that cohort fertility has fallen over time.
Otherwise, I follow the procedure used in the NFHS reports and calculate age specific 
fertility rates for five-year age groups using the number of births in months 1 to 36 
before the survey month (births in the survey month are not included) and the number
of women in each age group.%
\footnote{
See \citet{Croft2018} for more detail.
Because NFHS-1 and NFHS-2 did not interview all women about fertility, the number of women 
is based on the household rosters, assuming that never-married women have had no children.
Because of the low number of births to women aged 45 to 49 this age group was not
used for the hazard model estimations and, therefore, is also excluded for the fertility
rate.
The marital total fertility rate lead to unrealistically high numbers and is, therefore,
not presented \citep{Hoem2011,Laplante2015}.
}

@@

[TK things to cover:
- almost everybody marry, so these numbers should reflect cohort fertility relatively well
- what happens when women begin to join the labor force in larger numbers?
- better job oppportunities may be a direct way to affect sex selection since it makes 
it more expensive to have long spacing
]

[TK future research:
- health effects of multiple abortions on women and prior children
- better address selection effects for mortality estimations
- long-term outcomes for children
- labor force participation
- ]

The central question addressed here is the extent to which spacing patterns significantly 
changed as prenatal sex determination became available.
The underlying idea is that in the absence of sex selection son preference leads to shorter 
spacing after the birth of a girl than after the birth of a boy, whereas when prenatal sex 
selection is available son preference increases the spacing after the birth of a girl 
relative to after the birth of a boy.
I apply an empirical method that simultaneously accounts for spacing between births 
and the potential use of sex selection to four decades of data from India's NFHS.
